Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-13 submission/manuscript.Rmd
edit results 4

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have been conducted exclusively in males. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022]. We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.

Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4A-B). We compared the known functions of germline genes dysregulated in all samples (shared), only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), and only in males (XY only). Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic (meiotic cell cycle) and flagellar (cilium movement) genes, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion).

Germline genes dysregulated in both sexes were also enriched for meiotic ontologies (meiotic nuclear division), as well as egg-specific genes (female gamete generation). The majority of these shared germline DEGs had a greater log2 fold change from wild-type in females compared to males (Figure 4D-F). The increased number of germline genes and degree of dysregulation in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes[XXX]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 4G). Thus, while female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, it is likely independent of potential XCI defects.

>>>>

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have focused on their regulation in males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022]. We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.

Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B). We the compared the known functions of germline genes dysregulated only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared). Female-specific germline DEGs were enriched for meiotic (GO:XXX meiotic cell cycle) and flagellar (GO:XXX cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:XXX protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (meiotic nuclear division) and egg-specific functions (female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change from wild-type in females compared to males (Figure 4D). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to the loss of KDM5C-mediated suppression of germline genes. Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants, as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes instead lying on autosomes (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.
